Mapping mutational effects along the evolutionary landscape of HIV envelope

### 2018-02-26 paper/ResponseToReviews.tex

Of note, we also repeated the statistical analyses in Figure 7B and Figure 7C with the open conformation.
In both cases, the results were qualitatively similar for both conformations.
Thus, our initial conclusions about the solvent accessibility (Figure 7B) and clustering of shifted sites (Figure 7B) for the closed state also hold for the open state.

Next, we investigated substitution-shift co-localization in the post-fusion six-helix bundle.
Only $\sim$80 residues per Env monomer are resolved in available structures of the six-helix bundle.
Of these structures, the one with largest number of shifted sites (PDB: 1ENV) only had a total of four shifted sites and five substituted sites.
These numbers are too low to perform the same quantitative statistical tests from Figure 7 with the six-helix bundle.
But, we have added a new figure (Figure 7 - Figure supplement 2) that qualitatively analyzes substitution-shift co-localization.
We found that three of the four shifted sites cluster at one end of the helix, along with one of the five substitutions.
We wondered whether this cluster was unique to the six-helix bundle, but found that it was also present in the closed and open states as well (see the first two panels from the right for this figure).
Thus, although we agree that it was valuable to analyze the six-helix bundle in addition to the closed and open states, we did not observe any new patterns of co-localization that were not also present in the other states.

We have added the results of the new analyses described above to the first two paragraphs of the section in the Results entitled ``Structural and evolutionary properties of shifted sites", interleaved with the original text in these paragraphs.

@@

In total, we have expanded our structural analyses to include not only the closed state, but also the open state and the six-helix bundle.
This had made our analysis much more comprehensive.
Further, the agreement of results between the different structures, particularly the closed and open states, has substantially strengthened our initial conclusion about substitution-shift co-localization.

Note: In the process of making these revisions, we discovered a minor bug that affected our calculations of residue-residue distances. We have fixed that bug in the revised version. None of the qualitative results change (the trends that were previously significant are still significant, and the ones that previously were not significant are still not significant. However, some of the quantitative values in the box plots and the corresponding P-values have changed slightly.
}

3. Overall, it is found that the shifted sites tend to cluster together, but they are not necessarily the same as the sites with substitutions between the two Envs or ones that contact them. This analysis is not pursued in great detail, however, leaving the reasons why shifted sites tend to cluster together unexplained. Given that the authors also observe some evidence of entrenchment of substitutions between the Envs, is it possible that interactions between the shifted sites themselves are important? This question could be explored through a more detailed examination of the structure.

{\color{black}
This comment raises a very interesting question.
To answer it, we set about examining clusters of shifted sites in greater detail.
We found that two clusters occur within highly dynamic regions of Env.
One hypothesis as to why shifted sites cluster in these regions is that their highly dynamic nature allows long-range epistatic interactions to be readily propagated between shifted and substituted sites that are distant from one another.

The reviewers asked whether it is possible that the shifted sites interact and whether this could help explain the entrenchment we observe. If the above hypothesis is true, then shifted sites could interact by virtue of all being linked in a dynamic network. Though, because Env's dynamics are so complex, the exact mechanism of epistasis for entrenched residues is difficult to discern.

We have added a new supplemental figure (Figure 7 - figure supplement 3) that provides a fine-grained structural view of each of the above clusterss.
We have also added an entire paragraph detailing these observations at the end of the section in the Results called ``Structural and evolutionary properties of shifted sites".
Overall, these additions strengthen the manuscript by providing a better structural intuition for how long-range epistatic interactions might occur in Env.
}

### 2018-02-26 paper/manuscript.tex
Add response to major revision 3

\figsupp[\label{figsupp:shifts_on_six_helix_bundlel}Cluster of shifted and substituted sites in context of the post-fusion six-helix bundle of Env.]
{
The left-most panel shows the six-helix bundle of Env~\citep[PDB 1ENV;][]{weissenhorn1997atomic}, focusing on the central three helices, each of which is derived from a different Env monomer and which pack together to form the bundle's core.
There are $\sim$80 residues per monomer that are resolved in this structure, including four significantly shifted sites and five substituted sites.
Of these, three of the shifted sites (582, 583, and 587) and one substituted site (588) form a cluster at one end of the bundle.
We show the side chains for each of these sites colored according to whether the site has shifted (red) or substituted (yellow).
We wondered whether this cluster was unique to the six-helix bundle, but found that it is also present in roughly the same structural arrangement in both the closed pre-fusion conformation (center panel) and the open CD4-bound conformation (right panel; note that some of the surrounding residues in both the closed and open structures have been hidden for the sake of clarity).
As shown in \FIG{prefsdist}C, each of the shifted sites are more tolerant of mutations in BG505 than in BF520, indicating they are of greater structural and functional importance in BF520.
The two shifted sites that point inward towards the bundle's core (583 and 587) may be important for packing.
Based on these structures, it is difficult to discern if or how the substituted site may impact the preferences at the shifted sites nearby.
}
{\includegraphics[width=\textwidth]{figures/shifts_on_structure/six_helix_bundle_cluster.pdf}}

\figsupp[\label{figsupp:shifted_site_clusters}Clusters of shifted and substituted sites in context of different conformations of Env.]
{This figure gives a fine-grained view of three clusters of shifted sites in context of different conformations of Env.
In each structure, we only show side chains for shifted sites or other sites of interest, such as substituted sites.
Side chains are colored differently according to the key.
{\bf (A)}
This panel shows a cluster of four shifted sites at the apex of the closed pre-fusion conformation of Env~\citep[PDB 5FYL;][]{stewart2016trimeric}.
Each of these shifted sites has also substituted.
Further, this cluster is adjacent to other substituted sites, raising the possibility of epistasis among these sites.
This region of the structure involves packing between variable loops (V1/V2 and V3) and between these loops and an adjacent Env protomer.
This packing is known to control the conformational dynamics of these loops~\cite{guttman2014cd4,powell2017plasticity}.
Notably, a recent study found that mutations at two of the shifted sites (165 and 307) resulted in Env trimers with aberrant conformations~\cite{lee2017broadly}.
Thus, it is possible that this cluster of shifted sites is involved in controlling Env's dynamics.
Indeed, although each of these shifted sites prefers hydrophobic amino acids in both homologs, one homolog often has a higher preference for bulkier aromatic amino acids than the other.
This difference could make structural sense if these residues pack in different arrangements in the different homologs, possibly due to the substitutions in this region.
However, this hypothesis is difficult to test since there is not a high-resolution structure for BF520.
{\bf (B)}
This panel shows a cluster of shifted sites that are adjacent to a network of hydrophobic residues (blue/purple) that help mediate the conformational change between the closed pre-fusion state to the open CD4-bound state~\cite{ozorowski2017open}.
These sites are shown in context of both states, the latter of which is from~\citep[PDB 5VN3;][]{ozorowski2017open}.
Nearly one third of the shifted residues cluster within this structural region (9/30).
In contrast to panel (A), only a few of the shifted sites are adjacent to substitutions or have substituted themselves.
This trend is found in both the closed and open conformations, suggesting that the shifts may be primarily due to long-range epistatic interactions.
Such long-range interactions seem especially plausible in this region given that it is dynamic over a large swath or residues.
Indeed, one of the shifted sites (69) overlaps with the hydrophobic network, and mutations at another (64) have been shown to modulate the relative stability of the open and closed conformations~\cite{de2015immunogenicity}.
}
{\includegraphics[width=\textwidth]{figures/shifts_on_structure/clusters_on_structure.pdf}}
\end{figure}

What distinguishes the sites that have undergone significant shifts?
First, we analyzed the distribution of shifted sites in context of Env's three-dimensional structure.
Env's structure is highly conformationally dynamic and undergoes large changes upon receptor binding and membrane fusion.
In an effort to account for these dynamics, we examined multiple conformational states of Env, including: the closed pre-fusion state~\cite{julien2013crystal,pancera2014structure,stewart2016trimeric}, the open CD4-bound state~\cite{pancera2010structure,ozorowski2017open}, and the post-fusion six-helix bundle~\cite{weissenhorn1997atomic,chan1997core,tan1997atomic}.
\FIG{shiftfeatures}A shows the locations of the shifted sites on the crystal structure of Env in the closed pre-fusion state.
There is no visually obvious tendency for shifted sites to preferentially be on Env's surface or in its core, and statistical analysis of both the closed and open states of Env (\FIG{shiftfeatures}B) finds no association between a site's relative solvent accessibility and whether its amino-acid preferences have shifted.
However, \FIG{shiftfeatures}A does suggest that the sites of significant shifts tend to cluster in Env's structure, and a statistical analysis confirms that this is the case for both the closed and open states (\FIG{shiftfeatures}C).
Therefore, the factors that drive shifts in Env's mutational tolerance often affect physically interacting clusters of residues in a coordinated fashion.
Moreover, these factors are present in multiple conformations of Env.
We also investigated the distribution of shifted sites in context of the post-fusion six-helix bundle.
However, only $\sim$80 residues are resolved in this structure, and of these sites, only four have shifted amino-acid preferences.
This low number of sites does not lend itself to statistical analysis, but a qualitative analysis of these sites on the structure revealed that three of the four shifted sites cluster at one end of the helical bundle (\FIGSUPP[shiftfeatures]{shifts_on_six_helix_bundlel}).

@@

>>>>

\figsupp[\label{figsupp:shifts_on_six_helix_bundlel}Cluster of shifted sites in the post-fusion six-helix bundle of Env.]
{
The left-most panel shows the six-helix bundle of Env~\citep[PDB 1ENV;][]{weissenhorn1997atomic}, focusing on the central three helices, each of which is derived from a different Env monomer and which pack together to form the bundle's core.
The other three helices that make up this bundle are located in another part of the structure.
There are $\sim$80 residues per monomer that are resolved in this structure, including four significantly shifted sites and five substituted sites.
Of these, three of the shifted sites (582, 583, and 587) and one substituted site (588) form a cluster at one end of the bundle.
We show the side chains for each of these sites colored according to whether the site has shifted (red) or substituted (yellow).
We wondered whether this cluster was unique to the six-helix bundle, but found that it is also present in roughly the same structural arrangement in both the closed pre-fusion conformation (center panel) and the open CD4-bound conformation (right panel; note that some of the surrounding residues in both the closed and open structures have been hidden for the sake of clarity).
As shown in \FIG{prefsdist}C, each of the shifted sites are more tolerant of mutations in BG505 than in BF520, indicating they are of greater structural and functional importance in BF520.
The two shifted sites that point inward towards the bundle's core (583 and 587) may be important for packing.
Based on these structures, it is difficult to discern if or how the substituted site may impact the preferences at the shifted sites nearby.
}
{\includegraphics[width=\textwidth]{figures/shifts_on_structure/six_helix_bundle_cluster.pdf}}

\figsupp[\label{figsupp:dynamic_clusters}Clusters of shifted sites in highly dynamic regions of Env.]
{This figure gives a fine-grained view of two clusters of shifted sites, both of which reside in highly dynamic regions of Env.
In each structure, we only show side chains for shifted sites or other sites of interest, which include substituted sites and sites that are proposed to take part in a hydrophobic network that helps control Env's dynamics~\cite{ozorowski2017open}.
These side chains are colored differently according to the key.
{\bf (A)}
This panel shows a cluster of four shifted sites at the apex of the closed pre-fusion conformation of Env~\citep[PDB 5FYL;][]{stewart2016trimeric}, a region that is not resolved in the open CD4-bound conformation.
These sites are in Env's first/second variable loop (164 and 165) and its third variable loop (307 and 309), which pack against one another and against an adjacent Env protomer (the inter-protomer boundary is indicated by a dashed line).
Each of the shifted sites has substituted.
And these sites are in the immediate vicinity of other non-shifted, but substituted sites.
Thus, it seems likely that these shifts at least partially arise from short-range epistatic interactions within this cluster.
However, longer-range epistatic interactions with more distant sites also seem plausible given the highly dynamic nature of this region.
{\bf (B)}
This panel shows a cluster of shifted sites that are adjacent to a network of hydrophobic residues (blue/purple) that help mediate the conformational change between the closed pre-fusion state to the open CD4-bound state~\cite{ozorowski2017open}.
One of the shifted sites (purple) is in this network.
These sites are shown in context of both structural states, the latter of which is from~\citep[PDB 5VN3;][]{ozorowski2017open}.
Nearly one third of the shifted residues cluster within this structural region (9/30).
In contrast to panel (A), only a few of the shifted sites are adjacent to substitutions or have substituted themselves.
This trend is found in both the closed and open conformations, suggesting that the shifts may be primarily due to long-range epistatic interactions.
}
{\includegraphics[width=\textwidth]{figures/shifts_on_structure/clusters_on_structure.pdf}}
\end{figure}

What distinguishes the sites that have undergone significant shifts?
First, we analyzed the distribution of shifted sites in context of Env's three-dimensional structure.
Env's structure is highly conformationally dynamic and undergoes large changes upon receptor binding and membrane fusion.
In an effort to account for these dynamics, we examined multiple conformational states of Env, including: the closed pre-fusion state~\cite{julien2013crystal,pancera2014structure,stewart2016trimeric}, the open CD4-bound state~\cite{pancera2010structure,ozorowski2017open}, and the post-fusion six-helix bundle~\cite{weissenhorn1997atomic,chan1997core,tan1997atomic}.
\FIG{shiftfeatures}A shows the locations of the shifted sites on the crystal structure of Env in the closed pre-fusion state.
There is no visually obvious tendency for shifted sites to preferentially be on Env's surface or in its core, and statistical analysis of both the closed and open states of Env (\FIG{shiftfeatures}B) finds no association between a site's relative solvent accessibility and whether its amino-acid preferences have shifted.

However, \FIG{shiftfeatures}A does suggest that the sites of significant shifts tend to cluster in Env's structure, and a statistical analysis confirms that this is the case for both the closed and open states (\FIG{shiftfeatures}C).
Therefore, the factors that drive shifts in Env's mutational tolerance often affect physically interacting clusters of residues in a coordinated fashion.
Moreover, these factors are present in multiple conformations of Env.
We also investigated the distribution of shifted sites in context of the post-fusion six-helix bundle.
However, only $\sim$80 residues are resolved in this structure, and of these sites, only four have shifted amino-acid preferences.
This low number of sites does not lend itself to statistical analysis, but a qualitative analysis of these sites on the structure revealed that three of the four shifted sites cluster at one end of the helical bundle (\FIGSUPP[shiftfeatures]{shifts_on_six_helix_bundlel}).

@@

Indeed, many of the shifted sites cluster within regions of Env that are highly dynamic.
\FIGSUPP[shiftfeatures]{dynamic_clusters} shows the structural context of these clusters in finer detail.
One cluster is at the trimer apex where two of Env's variable loops pack against one another and against an adjacent protomer.
Upon CD4 binding, these loops become highly disordered~\cite{guttman2014cd4,ozorowski2017open}.
Mutations at two of the shifted sites in this cluster (165 and 307) have been shown to cause Env to assume aberrant conformations, suggesting that these sites can strongly modulate Env's dynamics~\cite{lee2017broadly}.
The other cluster of shifted sites is near a network of hydrophobic amino acids that has been proposed to help transmit the large-scale conformational change that takes place upon CD4 binding~\cite{ozorowski2017open}.
One of the shifted sites (69) overlaps with this network, and mutations at another (64) have been shown to strongly modulate the relative stability of the open and closed conformations~\cite{de2015immunogenicity}.
In total, these two clusters consist of nearly half (13/30) of the shifted sites.
One hypothesis as to why so many shifted sites cluster in these regions is that their dynamic nature allows long-range epistatic interactions to be readily propagated between substituted sites and distant shifted sites.
Indeed, as shown in \FIGSUPP[shiftfeatures]{dynamic_clusters}, these dynamic regions also contain several substituted sites.
However, due to the complexity of Env's dynamics, it is difficult to discern exactly how these interactions might occur.
